

```{r}
library(ggplot2)

N <- 30
sd <- 3
nor.mean <- 68.5
pop.mean <- 67
nor.zscore <- (nor.mean - pop.mean) / (sd / sqrt(N))

x <- seq(-4.5,4.5,0.01)
shade_x <- seq(nor.zscore,4.5,0.01)
shade_y <- dnorm(shade_x)
shade <- data.frame( rbind(c(nor.zscore,0), cbind(shade_x,shade_y), c(4.5,0)) )

ggplot() + 
    stat_function(aes(x=x), 
                  fun=dnorm, 
                  size=1, 
                  colour="blue") +
    ggtitle("Standardized sampling distribution\n(sample size N=30)") + 
    geom_hline(y=0, colour="darkgray") +
    geom_vline(x=0, linetype="dashed", colour="red", size=1) +
    geom_vline(x=nor.zscore, linetype="dashed", colour="orange", size=1) +
    geom_polygon(data = shade, aes(shade_x, shade_y), fill="yellow") +
    ylab("Probability density") + 
    xlab("Sample mean z-scores")  +
    scale_x_continuous(breaks=-4:4, labels=-4:4)

```

The area shaded in yellow is the probability of observing a sample mean z-score as extreme as the Norwegian sample mean.  As in the previous article, we can calculate the probability in R using the pnorm function:

```{r}
nor.zscore
1 - pnorm(nor.zscore)
```

Note that this is the exact same probability we got in the non-standardized significance test.  

## Recap



\begin{align*}
\overline{X} & \sim \operatorname{N}\left(\mu, \frac{\sigma^2}{N}\right )
\\[8pt]
\operatorname{E}[\overline{X}] & = \mu = 67in
\\[8pt]
\operatorname{Var}(\overline{X}) & = \frac{\sigma^2}{N} = \frac{3^2}{30}
\\[8pt]
\operatorname{SE}_{\overline{X}} & = \sqrt{\frac{\sigma^2}{N}} = \frac{\sigma}{\sqrt{N}} = \frac{3}{\sqrt{30}}
\\ \\
where,
\\[8pt]
\overline{X} & = \ the \ sample \ mean 
\end{align*}

